Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-08-18 paper.tex
Update on Overleaf.

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $k$ is reached. The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

@@

\comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse). Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1. En R pour un level plot}[(JR) oui tu as carrément raison, j'ai updaté les figures !]

@@

@@

\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]

\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}

\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}

\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}

\comment[MLT]{Mise à disposition des grilles générées? D'un échantillon?}

Ref Clem : \cite{alizadeh2016generating} génération de réseau spatiaux / données synthétiques

\paragraph{Comparing phase diagrams}

Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point. For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo? Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}. Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.

@@

With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
%%%%%%%%%%%%%
% Acknowledgements

>>>>

Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{Raimbault2014}. In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $k$ is reached. The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

@@

\comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse). Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.}[(JR) oui tu as carrément raison, j'ai updaté les figures !]\comment[CC]{magnifique! bien joué}

@@

\comment[cc]{Resultats Temporaires sur 1/3 des simu}
\begin{table*}[]
\centering
\caption{My caption}
\begin{adjustwidth}{-0.5cm}{-0.5cm}
\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.222 *** & -0.206 ***                       & -0.272 ***        & -0.288 ***        & -0.040 ***           & -0.086 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.426 ***  & 1.426 ***                        & 1.482 ***         & 1.482 ***         & 0.673 ***            & 0.673 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -1.158 *** & -1.158 ***                       & -1.179 ***        & -1.179 ***        & -0.545 ***           & -0.545 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.614 ***  & 0.614 ***                        & 0.486 ***         & 0.486 ***         & -0.037 ***           & -0.037 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.280 ***  & 0.280 ***                        & 0.094 ***         & 0.094 ***         & -0.090 ***           & -0.090 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.098 ***                        &                   & 0.070 ***         &                      & 0.002 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.117 ***                        &                   & 0.082 ***         &                      & -0.001 ***               \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.055 ***                       &                   & -0.024 ***        &                      & 0.026 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.270 ***                        &                   & 0.175 ***         &                      & \textit{-0.00}           \\ \hline
\textbf{R2 (\%)}                            & 34.6       & 38.5                             & 29.1              & 30.8              & 29.8                 & 30.0                     \\ 
\textbf{\# of observations (sim. runs)}     & 1,403,899  & {\color[HTML]{000000} 1,403,899} & 1,403,899         & 1,403,899         & 1,403,899            & 1,403,899                \\ 
\textbf{AIC}                                & -5939.02   & -91944.86                        & 246882.4          & 213145.1          & -2829106             & -2832773                 \\ \hline
\end{tabular}
\end{adjustwidth}
N.B. Moran's I applies to the minority population
\end{table*}

@@

We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled. We discuss them in this section.

\subsection{Limits}

\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point. For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo? Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}. Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.

\paragraph{Platform constrains and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space. For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements. Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space. This issue is part of the docking challenge \citep{Axtelletal1996} (i.e. checking if two models can produce the same results), but more generally, it involves a description of the model and its spatial requirements more detailed than what is currently the rule. \comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]

\subsection{Opportunities/extensions}

\paragraph{Reproducibility/Applicability.} \comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations. The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
We want foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow \comment[CC]{add link to github repo}. Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents.

\paragraph{An emancipation opportunity for social sciences.
\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} }

@@

After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration. We did so by implementing a spatial generator whose output is used as input for the simulation model. We applied this approach to two mythical ABMs: Schelling and Sugarscape. With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space. We think this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
%%%%%%%%%%%%%
% Acknowledgements
